Estimation of the Effect of a Continuous Treatment Variable

### 2025-11-28 writing/manuscript_CF_continuous.qmd
Changing figure label

Efficient nitrogen (N) management remains one of the most critical challenges in modern agricultural production. Nitrogen is essential for achieving high crop yields, yet excessive or poorly timed applications can lead to economic losses and environmental harm (<span style = "color: blue;"> add citations here </span>). The principle of applying the right amount of nitrogen at the right place and time requires understanding the spatial and temporal variability in crop nitrogen needs. This variability arises from complex interactions among soil nitrogen availability, soil and field characteristics, and weather conditions. Accurately identifying the optimal nitrogen rate for each site and season is therefore fundamental to improving both profitability and sustainability in crop production systems.

@@

```{r fig-trial-design, fig.cap = "Example trial design"}
g_td <- readRDS(here::here("writing/Figures/g_td.rds"))
```

@@

Taken together, @fig-rmse-hist and @fig-avg-performance illustrate clear performance patterns across methods and training sizes. CF delivers low and tightly concentrated RMSE distributions even with limited data—consistently maintaining strong accuracy and stability for 1–5 fields (mean RMSE ≈ 14.7, 13.1, and 12.1, respectively), with only modest additional gains at 10 and 20 fields (11.9–11.8). SO_ANN initially lags behind CF (18.4 at one field), but its mean RMSE decreases steadily as more fields are added, surpassing CF performance at 10 and 20 fields (11.2 and 8.5, respectively). This pattern underscores the neural network’s superior ability to exploit richer spatial and agronomic heterogeneity when sufficient training data are available.

By contrast, NO_ANN exhibits substantial instability in the single-field case, with a heavy right tail and much higher errors, but rapidly converges toward the leading methods by 5–10 fields (mean RMSE improving to around 14–13), highlighting its strong dependence on sample size. SO_RF improves modestly with more fields (mid-20s RMSE for 10–20 fields) but remains consistently less accurate than SO_ANN or CF. NO_RF maintains broad, right-shifted distributions and higher mean errors throughout (≈ 36–40), confirming its role as a weak baseline. Finally, DO_ANN performs poorly throughout, with both histograms and means showing persistently high errors (≈ 90+) and negligible benefit from additional data, reflecting <span style = "color: red;"> over-correction </span>   in the orthogonalization process.

::: {.column-margin}
<span style = "color: blue;"> It is not over-correction. It is fundamentally messed up as the original functions are no longer retained. </span>
:::
<!--end of column-margin-->

@@

model_names <- function(models) {
  map <- c(
    "Simple_ANN" = "No_ANN",
    "RANN"       = "DO_ANN",
    "HALFRANN"   = "SO_ANN",
    "HALFRRF"    = "SO_RF",
    "NO_RF"      = "NO_RF",
    "CF"         = "CF"
  )
  models_upper <- toupper(models)
  out <- map[models_upper]
  ifelse(is.na(out), models, out)
}

@@

# ggsave(
#   filename = here::here("Proposal_presentation/fig_rmse_hist.png"),
#   plot = p,
#   width = 12,
#   height = 7,
#   bg = "white"
# )

@@

rmse_df <- rmse_tbl_raw %>%
  dplyr::group_by(Model, `#field`, sim) %>%  
  dplyr::summarise(RMSE = dplyr::first(RMSE), .groups = "drop") %>%
  dplyr::mutate(
    # Field labels (auto from unique values)
    field_f = factor(
      `#field`,
      levels = sort(unique(`#field`)),
      labels = paste0(sort(unique(`#field`)),
                      ifelse(sort(unique(`#field`)) == 1, " field", " fields"))
    )
  )

# 3) summary stats 
rmse_summary <- rmse_df %>%
  dplyr::group_by(Model, field_f) %>%
  dplyr::summarise(
    n      = dplyr::n(),
    mean   = mean(RMSE, na.rm = TRUE),
    median = median(RMSE, na.rm = TRUE),
    sd     = sd(RMSE, na.rm = TRUE),
    se     = sd / sqrt(n),
    lower  = mean + qt(0.025, df = n - 1) * se,
    upper  = mean + qt(0.975, df = n - 1) * se,
    .groups = "drop"
  )
# 
p_mean <- ggplot(rmse_summary, aes(x = field_f, y = mean, group = Model, color = Model)) +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  geom_point(size = 3) +
  labs(x = "# fields", y = "Mean RMSE", color = "Model") +
  theme_bw() +
  theme(
    axis.title = element_text(size = 13),
    strip.text = element_text(face = "bold")
  )

@@

# ggsave(here::here("Proposal_presentation","model_improvement.png"), p_mean, width = 9, height = 5, dpi = 300)
```

@@

@fig-improvement reports each method’s percentage improvement in mean RMSE relative to the NO_RF baseline. CF consistently provides the strongest gains among the robust approaches when data are scarce—reducing error by ≈63% with a single field, and climbing to ≈67–68% once 5–20 fields are available. SO_ANN exhibits a similar improvement trend but grows even more dominant as more diverse data accumulate, achieving ≈53% improvement at 1 field, ≈63% by 5 fields, and reaching ≈68% at 10 fields and ≈77% at 20 fields, the highest improvement across all models. NO_ANN transitions from a large negative result when trained on just one field (substantially worse than baseline) to strong relative improvements exceeding 65% by 10–20 fields, reinforcing its reliance on adequate sample support. SO_RF offers moderate improvement (≈38–43%) that changes little with added fields. DO_ANN remains entirely uncompetitive across all sample sizes—showing worsening performance relative to NO_RF by more than 100–150%, indicating severe model misspecification.

::: {.column-margin}
<span style = "color: blue;"> All this is probably redundant. What are you adding beyond the previous paragraphs? </span>
:::
<!--end of column-margin-->

Together, these results emphasize that while CF is the most reliable option under limited data availability, SO_ANN ultimately yields the largest relative error reduction when multi-field datasets are accessible, highlighting its superior scalability with richer training information.

@@

# compare to which model?
baseline <- "NO_RF"

@@

# baseline per field 
baseline_by_field <- avg_by %>%
  dplyr::filter(Model == baseline) %>%
  dplyr::select(field_f, baseline_rmse = mean_rmse)

# join and compute % improvement
rel <- avg_by %>%
  dplyr::left_join(baseline_by_field, by = "field_f") %>%
  dplyr::filter(!is.na(baseline_rmse)) %>%
  dplyr::mutate(
    pct_impr = 100 * (baseline_rmse - mean_rmse) / baseline_rmse
    # or: pct_impr = (1 - mean_rmse / baseline_rmse) * 100
  ) %>%
  dplyr::filter(Model != baseline)

@@

<span style = "color: blue;"> I do not see the purpose of this much. Contrast CF, SO-NN, and the worst (all in the same panel) across the number of fields. </span>

@@

This study evaluates alternative causal-ML strategies for estimating heterogeneous yield response to a continuous nitrogen treatment in an OFPE setting. CF consistently attains low and tightly concentrated EONR errors with only 1–5 training fields, indicating strong small-sample stability (see @fig-rmse-hist and @fig-avg-performance). SO_ANN improves monotonically with additional fields and ultimately dominates once ≥10 fields are available, achieving the lowest mean RMSE overall (e.g., 11.2 at 10 fields and 8.5 at 20 fields). These complementary strengths that CF is preferred under data-constrained conditions, whereas SO_ANN becomes the method of choice as multi-field datasets accumulate (see @tbl-best_model_by_field_count).

>>>>

Efficient nitrogen (N) management remains one of the most critical challenges in modern agricultural production. Nitrogen is essential for achieving high crop yields, yet excessive or poorly timed applications can lead to economic losses and environmental harm (@fassa2022site). The principle of applying the right amount of nitrogen at the right place and time requires understanding the spatial and temporal variability in crop nitrogen needs. This variability arises from complex interactions among soil nitrogen availability, soil and field characteristics, and weather conditions. Accurately identifying the optimal nitrogen rate for each site and season is therefore fundamental to improving both profitability and sustainability in crop production systems.

@@

```{r fig-trial-design, fig.cap = "Example trial design"}
readRDS(here::here("writing/Figures/g_td.rds"))
```

@@

Taken together, @fig-rmse-hist and @fig-avg-performance illustrate clear performance patterns across methods and training sizes. CF delivers low and tightly concentrated RMSE distributions even with limited data—consistently maintaining strong accuracy and stability for 1–5 fields (mean RMSE ≈ 14.7, 13.1, and 12.1, respectively), with only modest additional gains at 10 and 20 fields (11.9–11.8). SO_ANN initially lags behind CF (18.4 at one field), but its mean RMSE decreases steadily as more fields are added, surpassing CF performance at 10 and 20 fields (11.2 and 8.5, respectively). This pattern hilightes the neural network’s ability to exploit richer spatial and agronomic heterogeneity when sufficient training data are available.

By contrast, NO_ANN exhibits substantial instability in the single-field case, with a heavy right tail and much higher errors, but rapidly converges toward the leading methods by 5–10 fields (mean RMSE improving to around 14–13), highlighting its strong dependence on sample size. SO_RF improves modestly with more fields (mid-20s RMSE for 10–20 fields) but remains consistently less accurate than SO_ANN or CF. NO_RF maintains broad, right-shifted distributions and higher mean errors throughout (≈ 36–40), confirming its role as a weak baseline. Finally, DO_ANN performs poorly throughout, with both histograms and means showing persistently high errors (≈ 90+) and negligible benefit from additional data, showing that the orthogonalization step breaks the original relationships in the data instead of preserving them.

@@

model_names <- function(models) {
  map <- c(
    "NO_ANN" = "NO_NN",
    "DO_ANN" = "DO_NN",
    "SO_ANN" = "SO_NN",
    "NO_RF"  = "NO_RF",
    "SO_RF"  = "SO_RF",
    "CF"     = "CF"
  )

  models_upper <- toupper(models)
  out <- map[models_upper]

  ifelse(is.na(out), models, out)
}

@@

# ggsave(
#   filename = here::here("presentation/proposal/fig_rmse_hist.png"),
#   plot = p,
#   width = 12,
#   height = 7,
#   bg = "white"
# )

@@

model_names <- function(models) {
  map <- c(
    "CF"     = "CF",
    "DO_ANN" = "DO_NN",
    "NO_ANN" = "NO_NN",
    "SO_ANN" = "SO_NN",
    "NO_RF"  = "NO_RF",
    "SO_RF"  = "SO_RF"
  )

  models_upper <- toupper(models)
  out <- map[models_upper]
  ifelse(is.na(out), models, out)
}

rmse_df <- rmse_tbl_raw %>%
  dplyr::group_by(Model, `#field`, sim) %>%
  dplyr::summarise(RMSE = dplyr::first(RMSE), .groups = "drop") %>%
  dplyr::mutate(
    Model_lab = model_names(Model),
    field_f = factor(
      `#field`,
      levels = sort(unique(`#field`)),
      labels = paste0(
        sort(unique(`#field`)),
        ifelse(sort(unique(`#field`)) == 1, " field", " fields")
      )
    )
  )

# 3) summary stats 
rmse_summary <- rmse_df %>%
  dplyr::group_by(Model_lab, field_f) %>%
  dplyr::summarise(
    n      = dplyr::n(),
    mean   = mean(RMSE, na.rm = TRUE),
    median = median(RMSE, na.rm = TRUE),
    sd     = sd(RMSE, na.rm = TRUE),
    se     = sd / sqrt(n),
    lower  = mean + qt(0.025, df = n - 1) * se,
    upper  = mean + qt(0.975, df = n - 1) * se,
    .groups = "drop"
  )

# 
p_mean <- ggplot(rmse_summary, aes(x = field_f, y = mean,
                                   group = Model_lab, color = Model_lab)) +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  geom_point(size = 3) +
  labs(
    x = "# fields",
    y = "Mean RMSE",
    color = "Model"
  ) +
  theme_bw() +
  theme(
    axis.title = element_text(size = 13),
    strip.text = element_text(face = "bold"),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 12)
  )

@@

# ggsave(here::here("presentation", "proposal", "model_improvement.png"), p_mean, width = 9, height = 5, dpi = 300)
```

@@

@fig-improvement compares the performance of CF and SO_ANN relative to the worst-performing model, DO_ANN, across different numbers of fields. Both methods deliver large improvements in RMSE at all sample sizes. With only one field, CF and SO_ANN reduce error by roughly 80–84%. As more fields become available, both approaches strengthen further: CF’s improvement rises to about 87% by 5–10 fields, while SO_ANN continues to gain with increasing data, reaching nearly 91% at 20 fields.

@@

# compare to which model?
baseline <- "DO_ANN"

@@

models_to_keep <- c("SO_ANN", "CF")

avg_by_filtered <- avg_by %>%
  dplyr::filter(Model %in% models_to_keep | Model == baseline)

# baseline per field 
baseline_by_field <- avg_by_filtered  %>%
  dplyr::filter(Model == baseline) %>%
  dplyr::select(field_f, baseline_rmse = mean_rmse)

# join and compute % improvement
rel <- avg_by_filtered  %>%
  dplyr::left_join(baseline_by_field, by = "field_f") %>%
  dplyr::filter(!is.na(baseline_rmse)) %>%
  dplyr::mutate(
    pct_impr = 100 * (baseline_rmse - mean_rmse) / baseline_rmse
    # or: pct_impr = (1 - mean_rmse / baseline_rmse) * 100
  ) %>%
  dplyr::filter(Model != baseline)

@@

@@

This study evaluates alternative causal-ML strategies for estimating heterogeneous yield response to a continuous nitrogen treatment in an OFPE setting. CF consistently attains low and tightly concentrated EONR errors with only 1–5 training fields, indicating strong small-sample stability (see @fig-rmse-hist and @fig-avg-performance). SO_ANN improves monotonically with additional fields and ultimately dominates once ≥10 fields are available, achieving the lowest mean RMSE overall (e.g., 11.2 at 10 fields and 8.5 at 20 fields). These complementary strengths that CF is preferred under data-constrained conditions, whereas SO_ANN becomes the method of choice as multi-field datasets accumulate.
